Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-07-30 paper.qmd
fixed country name for authors, labelled all code chunks

---
title: "Looking at Non-Linear Dimension Reductions as Models in the Data Space"
format: 
    jasa-pdf:
        keep-tex: true
    jasa-html: default
author:
  - name: Jayani P.G. Lakshika
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0002-6265-6481
    email: jayani.piyadigamage@monash.edu
    url: https://jayanilakshika.netlify.app/
  - name: Dianne Cook
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu 
    url: http://www.dicook.org/
  - name: Paul Harrison
    affiliations:
      - name: Monash University
        department: MGBP, BDInstitute
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0002-3980-268X
    email: 	paul.harrison@monash.edu
    url: 
  - name: Michael Lydeamore
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0001-6515-827X
    email: michael.lydeamore@monash.edu
    url: 
  - name: Thiyanga S. Talagala
    affiliations:
      - name: University of Sri Jayewardenepura
        department: Statistics
        address: Gangodawila
        city: Nugegoda 
        country: Sri Lanka
        postal-code: 10100
    orcid: 0000-0002-0656-9789
    email: ttalagala@sjp.ac.lk 
    url: https://thiyanga.netlify.app/
tbl-cap-location: bottom
abstract: |
  Non-linear dimension reduction (NLDR) techniques such as tSNE, and UMAP provide a low-dimensional representation of high-dimensional ($p\text{-}D$) data using non-linear transformation. The methods and parameter choices can create wildly different representations, making it difficult to decide which is best, or whether any or all are accurate or misleading. NLDR often exaggerates random patterns, sometimes due to the samples observed. But NLDR views have an important role in data analysis because, if done well, they provide a concise visual (and conceptual) summary of $p\text{-}D$ distributions. To help evaluate the NLDR we have developed an algorithm to show the $2\text{-}D$ NLDR model in the $p\text{-}D$ space, viewed with a tour. One can see if the model fits everywhere or better in some subspaces, or completely mismatches the data. It is used to evaluate which $2\text{-}D$ layout is the best representation of the $p\text{-}D$ distribution and see how different methods may have similar summaries or quirks.

@@

<!--UMAP applied for Scurve data-->
```{r}
umap_scurve <- read_rds(file = "data/s_curve/s_curve_umap.rds")

@@

```{r}
## Compute hexbin parameters
num_bins_x_scurve <- 7

@@

```{r}
hex_grid_nonempty <- hex_grid |>
  filter(hex_poly_id %in% df_bin_centroids$hexID)

@@

```{r}
## Triangulate bin centroids
tr1_object_scurve <- tri_bin_centroids(
  df_bin_centroids, x = "c_x", y = "c_y")
tr_from_to_df_scurve <- gen_edges(
  tri_object = tr1_object_scurve)

@@

```{r}
# Code to draw illustration for notation
start_pt <- all_centroids_df |> 
  filter(hexID == 1)
d_rect <- tibble(x1min = 0, 
                 x1max = 1,
                 x2min = 0,
                 x2max = r2)

@@

```{r}
## hexagon binning to have regular hexagons
hb_obj_scurve1 <- hex_binning(
  data = umap_scurve_scaled, 
  bin1 = 7, 
  r2 = r2,
  q = 0.12)

@@

```{r}
## hexagon binning to have regular hexagons
hb_obj_scurve2 <- hex_binning(
  data = umap_scurve_scaled, 
  bin1 = 11, 
  r2 = r2,
  q = 0.07)

@@

```{r}
## hexagon binning to have regular hexagons
hb_obj_scurve3 <- hex_binning(
  data = umap_scurve_scaled, 
  bin1 = 17, 
  r2 = r2,
  q = 0.05)

@@

```{r}
training_data_scurve <- read_rds("data/s_curve/s_curve_training.rds")
```

<!--To generate errors for different total number of bins-->
```{r}
## To initialize number of bins along the x-axis
bin1_vec_scurve <- 2:19 #sqrt(NROW(training_data_scurve)/2)
buffer_vec <- seq(0.05, 0.1, by = 0.01)

@@

```{r}
## To initialize effective bins along x
effective_bin1_scurve <- 11

@@

```{r}
cell_count_scurve <- ggplot(df_bin_centroids_scurve, 
                            aes(x = reorder(as.factor(hexID), 
                                            -std_counts), 
                                y = std_counts)) + 
  geom_quasirandom(
    size = 1,
    alpha = 0.5
  ) + 
  geom_hline(yintercept = benchmark1, color = "black", linewidth=0.5, alpha = 0.5) +
  xlab("hexagon id (re-ordered)") + 
  ylab("standardized bin count") +
  ggtitle("(b)") +
  theme(aspect.ratio = 0.75,
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12, angle = 90),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 12),
        plot.title = element_text(size = 12))
```

@@

<!--To generate errors for different total number of bins-->
```{r}
## To initialize benchmark values to remove low density hexagons
benchmark_rm_hex_vec <- append(seq(0, 0.99, by=0.1), c(benchmark1, 0.99))

@@

<!--Using first quantile as the benchmark value-->
```{r}
hex_grid_with_counts_s_curve2 <- full_join(
  hex_grid2, 
  df_bin_centroids_scurve |> select(hexID, std_counts), 
  by = c("hex_poly_id" = "hexID"))

@@

<!--Observing neighboring bins as well-->
```{r}
## First define low density bins using first quantile
df_bin_centroids_low <- df_bin_centroids_scurve |> 
  filter(std_counts <= benchmark1)

@@

<!--distribution of distance along with the default benchmark-->
```{r}
## Triangulate bin centroids
tr1_object_scurve <- tri_bin_centroids(
  df_bin_centroids_scurve_keep, x = "c_x", y = "c_y")
tr_from_to_df_scurve <- gen_edges(
  tri_object = tr1_object_scurve)

@@

<!-- UMAP layout with author's suggested parameter choice-->
```{r}
## Import data
training_data_pbmc <- read_rds("data/pbmc3k/pbmc_pca_50.rds")
training_data_pbmc <- training_data_pbmc[, 1:9] |>
  mutate(ID = 1:NROW(training_data_pbmc))

@@

<!--Fit the best model for author suggestion and compute error-->
```{r}
## Compute hexbin parameters
num_bins_x_pbmc <- 30
lim1 <- pbmc_scaled_obj$lim1
lim2 <- pbmc_scaled_obj$lim2
r2_pbmc <- diff(lim2)/diff(lim1)

@@

```{r}
## To initialize number of bins along the x-axis
bin1_vec_pbmc <- 2:36 #sqrt(NROW(training_data_pbmc)/2)

@@

```{r}
error_pbmc <- bind_rows(error_pbmc_umap, 
                        error_pbmc_umap2,
                        error_pbmc_umap3,
                        error_pbmc_tsne,
                        error_pbmc_tsne2,
                        error_pbmc_phate,
                        error_pbmc_trimap,
                        error_pbmc_pacmap)
```

```{r}
# read the png file from device 
tsne_5_img <- readPNG("figures/pbmc3k/tsne_5.png", native = TRUE) 
phate_5_img <- readPNG("figures/pbmc3k/phate_5.png", native = TRUE) 
umap_img <- readPNG("figures/pbmc3k/author_view.png", native = TRUE) 
tsne_30_img <- readPNG("figures/pbmc3k/tsne_30.png", native = TRUE)

@@

```{r}
tsne_pbmc <- read_rds("data/pbmc3k/pbmc_tsne_30.rds")
pbmc_scaled_obj <- gen_scaled_data(
  data = tsne_pbmc)
tsne_pbmc_scaled <- pbmc_scaled_obj$scaled_nldr

@@

<!--Fit the best model and compute error-->
```{r}
## Compute hexbin parameters
num_bins_x_pbmc <- 15
lim1 <- pbmc_scaled_obj$lim1
lim2 <- pbmc_scaled_obj$lim2
r2_pbmc <- diff(lim2)/diff(lim1)

@@

<!-- PaCMAP layout with author's suggested parameter choice-->
```{r}
## Import data
training_data_mnist <- read_rds("data/mnist/mnist_10_pcs_of_digit_1.rds")
training_data_mnist <- training_data_mnist |>
  mutate(ID = 1:NROW(training_data_mnist))

@@

<!-- Fit the model and compute error-->
```{r}
## Compute hexbin parameters
num_bins_x_mnist <- 22
lim1 <- mnist_scaled_obj$lim1
lim2 <- mnist_scaled_obj$lim2
r2_mnist <- diff(lim2)/diff(lim1)

>>>>

---
title: "Looking at Non-Linear Dimension Reductions as Models in the Data Space"
format: 
    jasa-pdf:
        keep-tex: true
    jasa-html: default
author:
  - name: Jayani P.G. Lakshika
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-6265-6481
    email: jayani.piyadigamage@monash.edu
    url: https://jayanilakshika.netlify.app/
  - name: Dianne Cook
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu 
    url: http://www.dicook.org/
  - name: Paul Harrison
    affiliations:
      - name: Monash University
        department: MGBP, BDInstitute
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3980-268X
    email: 	paul.harrison@monash.edu
    url: 
  - name: Michael Lydeamore
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0001-6515-827X
    email: michael.lydeamore@monash.edu
    url: 
  - name: Thiyanga S. Talagala
    affiliations:
      - name: University of Sri Jayewardenepura
        department: Statistics
        address: Gangodawila
        city: Nugegoda 
        country: Sri Lanka
        postal-code: 10100
    orcid: 0000-0002-0656-9789
    email: ttalagala@sjp.ac.lk 
    url: https://thiyanga.netlify.app/
tbl-cap-location: bottom
abstract: |
  Non-linear dimension reduction (NLDR) techniques such as tSNE, and UMAP provide a low-dimensional representation of high-dimensional ($p\text{-}D$) data using non-linear transformation. The methods and parameter choices can create wildly different representations, making it difficult to decide which is best, or whether any or all are accurate or misleading. NLDR often exaggerates random patterns, sometimes due to the samples observed. But NLDR views have an important role in data analysis because, if done well, they provide a concise visual (and conceptual) summary of $p\text{-}D$ distributions. To help evaluate the NLDR we have developed an algorithm to show the $2\text{-}D$ NLDR model in the $p\text{-}D$ space, viewed with a tour. One can see if the model fits everywhere or better in some subspaces, or completely mismatches the data. It is used to evaluate which $2\text{-}D$ layout is the best representation of the $p\text{-}D$ distribution and see how different methods may have similar summaries or quirks.

@@

<!--UMAP applied for Scurve data-->
```{r}
#| label: umap-scurve
umap_scurve <- read_rds(file = "data/s_curve/s_curve_umap.rds")

@@

```{r}
#| label: hexbin-scurve
## Compute hexbin parameters
num_bins_x_scurve <- 7

@@

```{r}
#| label: empty-bin-scurve
hex_grid_nonempty <- hex_grid |>
  filter(hex_poly_id %in% df_bin_centroids$hexID)

@@

```{r}
#| label: triangulate-scurve
## Triangulate bin centroids
tr1_object_scurve <- tri_bin_centroids(
  df_bin_centroids, x = "c_x", y = "c_y")
tr_from_to_df_scurve <- gen_edges(
  tri_object = tr1_object_scurve)

@@

```{r}
#| label: code-illustration
# Code to draw illustration for notation
start_pt <- all_centroids_df |> 
  filter(hexID == 1)
d_rect <- tibble(x1min = 0, 
                 x1max = 1,
                 x2min = 0,
                 x2max = r2)

@@

```{r}
#| label: hexbin-regular-scurve1
## hexagon binning to have regular hexagons
hb_obj_scurve1 <- hex_binning(
  data = umap_scurve_scaled, 
  bin1 = 7, 
  r2 = r2,
  q = 0.12)

@@

```{r}
#| label: hexbin-regular-scurve2
## hexagon binning to have regular hexagons
hb_obj_scurve2 <- hex_binning(
  data = umap_scurve_scaled, 
  bin1 = 11, 
  r2 = r2,
  q = 0.07)

@@

```{r}
#| label: hexbin-regular-scurve3
## hexagon binning to have regular hexagons
hb_obj_scurve3 <- hex_binning(
  data = umap_scurve_scaled, 
  bin1 = 17, 
  r2 = r2,
  q = 0.05)

@@

```{r}
#| label: scurve-training
training_data_scurve <- read_rds("data/s_curve/s_curve_training.rds")
```

<!--To generate errors for different total number of bins-->
```{r}
#| label: errors-scurve
## To initialize number of bins along the x-axis
bin1_vec_scurve <- 2:19 #sqrt(NROW(training_data_scurve)/2)
buffer_vec <- seq(0.05, 0.1, by = 0.01)

@@

```{r}
#| label: distribution-scurve
## To initialize effective bins along x
effective_bin1_scurve <- 11

@@

```{r}
#| label: bin-counts-scurve
cell_count_scurve <- ggplot(df_bin_centroids_scurve, 
                            aes(x = reorder(as.factor(hexID), 
                                            -std_counts), 
                                y = std_counts)) + 
  geom_quasirandom(
    size = 1,
    alpha = 0.5
  ) + 
  geom_hline(yintercept = benchmark1, color = "black", linewidth=0.5, alpha = 0.5) +
  xlab("hexagon id (re-ordered)") + 
  ylab("standardized bin count") +
  ggtitle("(b)") +
  theme(aspect.ratio = 0.75,
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12, angle = 90),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 12),
        plot.title = element_text(size = 12))
```

@@

<!--To generate errors for different total number of bins-->
```{r}
#| label: bins-scurve
## To initialize benchmark values to remove low density hexagons
benchmark_rm_hex_vec <- append(seq(0, 0.99, by=0.1), c(benchmark1, 0.99))

@@

<!--Using first quantile as the benchmark value-->
```{r}
#| label: bin-benchmark-scurve
hex_grid_with_counts_s_curve2 <- full_join(
  hex_grid2, 
  df_bin_centroids_scurve |> select(hexID, std_counts), 
  by = c("hex_poly_id" = "hexID"))

@@

<!--Observing neighboring bins as well-->
```{r}
#| label: bin-neighbours
## First define low density bins using first quantile
df_bin_centroids_low <- df_bin_centroids_scurve |> 
  filter(std_counts <= benchmark1)

@@

<!--distribution of distance along with the default benchmark-->
```{r}
#| label: triangulate-scurve2
## Triangulate bin centroids
tr1_object_scurve <- tri_bin_centroids(
  df_bin_centroids_scurve_keep, x = "c_x", y = "c_y")
tr_from_to_df_scurve <- gen_edges(
  tri_object = tr1_object_scurve)

@@

<!-- UMAP layout with author's suggested parameter choice-->
```{r}
#| label: published-pbmc
## Import data
training_data_pbmc <- read_rds("data/pbmc3k/pbmc_pca_50.rds")
training_data_pbmc <- training_data_pbmc[, 1:9] |>
  mutate(ID = 1:NROW(training_data_pbmc))

@@

<!--Fit the best model for author suggestion and compute error-->
```{r}
#| label: hexbin-pbmc
## Compute hexbin parameters
num_bins_x_pbmc <- 30
lim1 <- pbmc_scaled_obj$lim1
lim2 <- pbmc_scaled_obj$lim2
r2_pbmc <- diff(lim2)/diff(lim1)

@@

```{r}
#| label: initial-bins-pbmc
## To initialize number of bins along the x-axis
bin1_vec_pbmc <- 2:36 #sqrt(NROW(training_data_pbmc)/2)

@@

```{r}
#| label: combine-data-pbmc
error_pbmc <- bind_rows(error_pbmc_umap, 
                        error_pbmc_umap2,
                        error_pbmc_umap3,
                        error_pbmc_tsne,
                        error_pbmc_tsne2,
                        error_pbmc_phate,
                        error_pbmc_trimap,
                        error_pbmc_pacmap)
```

```{r}
#| label: tsne-pbmc
# read the png file from device 
tsne_5_img <- readPNG("figures/pbmc3k/tsne_5.png", native = TRUE) 
phate_5_img <- readPNG("figures/pbmc3k/phate_5.png", native = TRUE) 
umap_img <- readPNG("figures/pbmc3k/author_view.png", native = TRUE) 
tsne_30_img <- readPNG("figures/pbmc3k/tsne_30.png", native = TRUE)

@@

```{r}
#| label: tsne-pbmc-read
tsne_pbmc <- read_rds("data/pbmc3k/pbmc_tsne_30.rds")
pbmc_scaled_obj <- gen_scaled_data(
  data = tsne_pbmc)
tsne_pbmc_scaled <- pbmc_scaled_obj$scaled_nldr

@@

<!--Fit the best model and compute error-->
```{r}
#| label: num-bins-pbmc
## Compute hexbin parameters
num_bins_x_pbmc <- 15
lim1 <- pbmc_scaled_obj$lim1
lim2 <- pbmc_scaled_obj$lim2
r2_pbmc <- diff(lim2)/diff(lim1)

@@

<!-- PaCMAP layout with author's suggested parameter choice-->
```{r}
#| label: read-mnist
## Import data
training_data_mnist <- read_rds("data/mnist/mnist_10_pcs_of_digit_1.rds")
training_data_mnist <- training_data_mnist |>
  mutate(ID = 1:NROW(training_data_mnist))

@@

<!-- Fit the model and compute error-->
```{r}
#| label: hexbin-mnist
## Compute hexbin parameters
num_bins_x_mnist <- 22
lim1 <- mnist_scaled_obj$lim1
lim2 <- mnist_scaled_obj$lim2
r2_mnist <- diff(lim2)/diff(lim1)
